\documentclass[a4paper, oneside]{book}
\usepackage{jipkg}
\usepackage{multirow}

%few ipa symbols
\newcommand{\ipaE}{ε}
\newcommand{\ipaET}{ɛ̃}
\newcommand{\ipaO}{ɔ}
\newcommand{\ipaAT}{ɑ̃}
\newcommand{\ipaS}{ʃ}
\newcommand{\ipaZ}{ʒ}
\newcommand{\ipaR}{ʁ}

%glossary
\usepackage[toc]{glossaries}
\newcommand{\newgle}[2]{\newglossaryentry{#1}{name=#1, description={#2}}}
\glossarystyle{listgroup}
\makeglossaries

%gloss and examples
\usepackage{gb4e}

\title{Deẽreẽ\\Grammar of a fantasy constructed language}
\author{Jirsad \textsc{Ïoh}}
\date{2016 — \today}
\jihypersetup{Deẽreẽ — A fantasy conlang}{Jirsad ÏOH}

\begin{document}
\selectlanguage{english}
\frontmatter
\maketitle
\tableofcontents
\jichap{Introduction}
\begin{exe}
\ex\label{ex:intro-cat}
\gll mosoj’a fãssa eat a\\
cat-3S wizard 3S.INDEF 3S\\
\trans ‘The cat is a wizard’
\end{exe}

In example \ref{ex:intro-cat}, we see that blah blah blah…
\todo{fig:glossing abbreviations}

\mainmatter
%-----
\jichap{Phonology}\label{chap:phonology}
\input{phonology/phonology}

%-----
\jichap{Morphology}\label{chap:morphology}
This chapter, morphology, aims to describe \emph{what the words look like}. This means describing
the nouns, adjectives, adverbs, verbs in their basic forms. Details though will be about person
suffixes, Deẽreẽ’s equivalent for articles.

\input{morphology/gender}
\input{morphology/number}
\input{morphology/personSuffixes}

\jisec{Adjective}
\todo{before: all, no suffix}

\todo{after: one only, 3S suffix}
\jisec{Adpositions}
\todo{post-positions a lot}

\todo{pre-positions with stuff.}

%-----
\jichap{Derivational Morphology}\label{chap:derivational-morphology}
The name ‘derivational morphology’ refers to all the mechanisms of word derivation, that is, how to
make a word out of another. These rules apply to several kinds of words, and are hereafter grouped
by part of speech they derive.

\jisec{Noun derivations}
\jisec{Verb derivations}
\todo{1t1 --> 2lf: tool, that with which X is done\\trasata: to give --> trasãlf: generosity}

\todo{1t1 --> 31n: result, that which X does\\torete: to say --> toruen: speech}

%-----
\jichap{Conjugation}\label{chap:conjugation}
\jisec{Tense}
\jissec{Present}
\todo{et}
\jissec{Past}
\todo{ẽiet}
\jissec{Future}
\todo{et trã}
\jisec{Mood}
\jissec{Indicative}
\todo{et}
\jissec{Participle}
\todo{ej}
\jissec{Interrogative}
\todo{Vf es}
\jissec{Subjunctive}
\todo{Vf es}
\jissec{Imperative}
\todo{Vf a-V-et}
\jisec{Aspect}
\jissec{Perfect/Imperfect}
\todo{drej}
\jissec{Cessative/Inchoative}
\todo{drẽiej/drej trã}
\jisec{The Copula}
\todo{first suffix repetition}

Mosoj’a fãssa eat’a. => the cat is a wizard.

%-----
\jichap{Syntax}\label{chap:syntax}
\jisec{Noun phrase}
\jisec{Sentence}
\jissec{Basic order}
\jissec{Questions}
\jissec{Imperatives}
\jisec{Subclauses}
\todo{Vf}
\jisec{Existentials}
\todo{at}

%-----
\jichap{Pragmatics}\label{chap:pragmatics}
\jisec{Sayings}
\jisec{Samples}
\input{pragmatics/samples/theKingOfAllSnakes}
\newgle{naklë}{snake}

%-----
\backmatter
%\input{appendices/swadesh}
\clearpage\glsaddall
\begin{multicols}{2}\printglossaries\end{multicols}
\listoffigures\addcontentsline{toc}{chapter}{\listfigurename}
\listoftables\addcontentsline{toc}{chapter}{\listtablename}
\listoftodos
\end{document}
